A Simplified Plackett-Luce Likelihood Function for Rank Orders

### 2025-11-25 inst/hankin-exponential_BT.tex
instructions for auxiliary files not under version control
more writing

\documentclass[article]{ajs}
\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
\RequirePackage{graphicx}%% uncomment this for including figures

@@

\noindent where the strengths are normalised so $\sum p_i=1$.
However, estimation of the strengths is somewhat difficult if $n$ is
even moderately large: even with as few as 20 competitors following
Zipf's law~\citep{zipf1949} [a typical assumption for Bradley--Terry
  strengths] and say 20 independent complete rank observations,
\cite{hankin2017_rmd} shows that maximum likelihood strengths
identify the correct ranking with a probability of essentially zero.
The situation is worse in cases such as formula 1 motor racing
considered below, which requires us to consider ${\mathcal O}(100)$
competitors.

@@

We consider the smallest non-trivial cases $n=3$ and $n=4$.  For $n=3$
we have Plackett-Luce strengths $\alpha x,\alpha x^2,\alpha x^3$ [the
  extra factor of $x$ allows us to identify the $r$ in ``$\alpha x^r$''
  with the rank of the competitor, the best being rank 1].  Suppose
the focal competitor came first, the likelihood function would be

\begin{equation}
  \mathcal{L}(x,r)=\frac{x^{r-1}}{1+x+x^2}\qquad r=1,2,3
\end{equation}

If the focal competitor came second we would have two Plackett-Luce
probabilities to add, corresponding to different order statistics.
$\alpha x^i$ or $\alpha x^j$ coming first where $i,j\neq r$.  For
example, if $r=1$ we would have two finishing orders that are
consistent with the observation: $2\succ 1\succ 3$ and $3\succ 1\succ
2$.  These are disjoint so the likelihood will be proportional to

@@

We leverage the {\tt hyper2} R package~\cite{hankin2017_rnw} which
includes extensive numerical functionality to simulate Plackett-Luce
situations, and in this case function {\tt rrace()} simulates a race
between competitors of known Plackett-Luce strengths.  To obtain a
likelihood function for a specific value of $(x,s)$ at a particular
observed rank $r$, we run repeated {\em in-silico} races between
competitors of Plackett-Luce strengths $x^1,x^2,\ldots x^n$, and
observe the frequency with which competitor $\lceil ns\rceil$ places
$r$.  For example, consider $x=0.1, r=3$.  We might simulate this as
follows:

@@

Above we see 15 independent Plackett-Luce races; for example the first
one indicates that competitor 2 came first, 1 came second, 4 came
third, and so on.  Looking at the fourth column [headed \verb+c4+,
  ``came fourth''], we see that

@@

The maximum likelihood estimate for $s$ would therefore be $4/9\simeq
0.44$, conditional on $x=0.1$.  To estimate $x$ we would need repeated
observations, and here we note that small values of $x$ lead to
underdispersed results.

>>>>

\documentclass[article]{ajs}  % need to download ajs.cls, oesg.png, ajs.bst from the Austrian Journal of Statistics github repo
\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
\RequirePackage{graphicx}%% uncomment this for including figures

@@

\noindent where the strengths are normalised so $\sum p_i=1$.
However, estimation of the strengths is somewhat difficult if $n$ is
even moderately large: even with as few as 20 competitors following
Zipf's law~\citep{zipf1949} [a typical assumption for Bradley-Terry
  strengths] and say 20 independent complete rank observations,
\cite{hankin2017_rmd} shows that maximum likelihood strengths
identify the correct ranking with a probability of essentially zero.
The situation is worse in cases such as formula 1 motor racing
considered below, which requires us to consider ${\mathcal O}(100)$
competitors.

In many situations, we are not interested in the $n-1$-dimensional
simplex of BT strengths, viz $\left\lbrace(p_1,\ldots,p_n)\colon
p_i\geqslant 0,\sum p_i=1\right\rbrace$, but instead wish to make
inferences about one particular entity.  For example, we might be
interested in whether Mercedes's apparent decline in performance in
the 2023-4 season could be due to chance.  Further examples are given
in previous analysis by the same author \citep{hankin2010,
  hankin2017_rmd, hankin2020, hankin2023_formula1points}.  In these
cases, we are interested in one of the competitors and assess a number
of hypotheses about that competitor; the strengths of the others are
nuisance parameters.  We designate this entity as the {\em focal
 competitor}.

If we have $n$ competitors, optimizing the Bradley--Terry model over
$n-1$ degrees of freedom is, as noted above, a formidable task even
for moderate $n$; this suggests that some form of dimension-reducing
parametrization might be useful.  One option might be to assign the
focal competitor a BT strength proportional to $a\in [0,1]$ and
$b=1-a$ to each of $n-1$ competitors; the normalization constant would
be $n+(n-1)a$.  However, this option does not seem to be
realistic~\citep{kuenzer2025}.

The next natural suggestion would be to have $n$ competitors with
strengths $\beta, \beta x,\ldots \beta x^{n-1}$ with $0 < x\leqslant
1$.  The unit sum constraint gives $\beta=\frac{1-x}{1-x^n}$ but this
constant of proportionality cancels out in Bradley-Terry likelihoods.
The focal competitor has strength $\beta x^{r}$ for some (unknown) $r$
with $1\leqslant r\leqslant n$.  Thus $r=1$ means that the focal
competitor is the strongest among the $n$, $r=2$ means he is the
second, and so on.

\subsection{Theoretical results}

A Plackett-Luce observation is of the form $p_{(1)}\succ
p_{(2)}\succ\cdots\succ p_{(n)}$, where $p_{(i)}$ denotes the
competitor in place $i $ for $1\leqslant i\leqslant n$.  If competitor
$j$ has strength $\beta x^{j}$, then the probability of observing
$p_1\succ p_2\succ\cdots\succ p_n$ will be, according to
equation~\ref{plackettluce},

\begin{equation}
f(x)=\frac{1}{1+\cdots+x^{n-1}}\cdot
\frac{1}{1+\cdots+x^{n-2}}\cdots
\frac{1}{1+x^{2}}\cdot\frac{1}{1+x}=
\frac{1}{[n]_x!}
\end{equation}

(here $[n]_x!$ is the $q$-factorial).  We see that
$f(1)=\frac{1}{n!}$, $f'(x)>0$ and $f(0)=1$; so small $x$ corresponds
to a high probability of observing the ``natural order'' in which
competitor $j$ places $j$-th, $1\leqslant j\leqslant n$.  Further, we
observe that $f(x)=1-(n-1)x + {\mathcal O}(x^2)$.  It is interesting
to note that the probability of observing the reverse order, that is
$p_n\succ p_{n-1}\succ\cdots\succ p_1$, is $f(1/x)$ for $x>0$.

@@

We consider the smallest non-trivial cases $n=3$ and $n=4$.  For $n=3$
we have Plackett-Luce strengths $\alpha x,\alpha x^2,\alpha x^3$ [the
  extra factor of $x$ allows us to identify the $r$ in ``$\alpha
  x^r$'' with the nominal rank of the competitor, the best being rank
  1].  Suppose the focal competitor came first, the likelihood
function would be

\begin{equation}
  \mathcal{L}(x,r)=\frac{x^{r-1}}{1+x+x^2}\qquad r=1,2,3.
\end{equation}

If the focal competitor came second we would have two Plackett-Luce
probabilities to add, corresponding to different order statistics.
For example, if $r=1$ we would have two finishing orders that are
consistent with the observation: $2\succ 1\succ 3$ and $3\succ 1\succ
2$.  These are disjoint so the likelihood will be proportional to

@@

We leverage the {\tt hyper2} R package~\citep{hankin2017_rmd} which
includes extensive numerical functionality to simulate Plackett-Luce
situations, and in this case function {\tt rrace()} simulates a race
between competitors of known Plackett-Luce strengths.  To obtain a
likelihood function for a specific value of $(x,s)$ at a particular
observed rank $r$, we run repeated {\em in-silico} races between
competitors of Plackett-Luce strengths $x^1,x^2,\ldots x^n$, and
observe the frequency with which competitor $\lceil ns\rceil$ places
$r$.  For example, consider $x=0.1, r=3$.  We might simulate this as
follows:

@@

Above we see a bespoke {\tt hyper2} object, of class {\tt ranktable},
designed to display rank data for use with Bradley-Terry likelihoods.
We see 15 independent Plackett-Luce races; for example the first one
indicates that competitor 2 came first, 1 came second, 4 came third,
and so on.  Looking at the fourth column [headed \verb+c4+, ``came
  fourth''], we see that

@@

The maximum likelihood estimate for $s$ would therefore be $4/9\simeq
0.44$, conditional on $x=0.1$.  To estimate $x$ we would need repeated
observations, and here we note that small values of $x$ lead to
underdispersed results, for the focal competitor would come $r$-th
with probability $\geqslant x^1+x^2+\ldots x^r$
